---
title: "Homework"
author: "YOUR NAME"
output: html_document
date: "2024-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Load libraries

```{r}
library(tidyverse)
```

#### Read in data 

```{r}
#Load data
surveys <- read_csv("diversity.csv")
```

### Tidyverse Day 1

#### Q1: What is the class of the object surveys? 

```{r}
#Q1

```


#### Q2: How many rows and how many columns are in the object?

Rows: 
Columns:

#### Q2: Create a data.frame (surveys_200) containing only the data in row 200 of the surveys dataset.

```{r}
#Q2

```

#### Q3: Create a data.fram (surveys_last) containing only the data in row 200 of the surveys dataset using nrow() instead of the row number. 

```{r}
#Q3

```

#### Q4: Change the columns taxa and genus in the surveys data frame into a factor.

```{r}
#Q4

```

#### Q5: Using the functions you learned before

##### Q5 A: How many rabbits were observed? 

```{r}
#Q5A

```


##### Q5 B: How many different genera are in the genus column?

```{r}
#Q5B

```

#### Q6: Create new variable 'sex' by converting surveys$sex to a factor. Use addNA to add a factor level to 'sex' variable. Using the 'sex' variable, rename "NA", "F", and "M" factors to "undetermined", "female", and "male" respectively. Plot sex variable. 

```{r}
#Q6

```

#### Q7: Recreate the barplot such that "undetermined" is first (before "female").

```{r}
#Q7

```

#### Q8: Using pipes, subset the surveys data to include animals collected before 1995 and retain only the columns year, sex, and weight. View the tail() of the data.frame.

```{r}
#Q8

```

### Tidyverse Day 2

#### Q9: Create a new data frame from the surveys data that meets the following criteria:

##### Contains only the species_id column and a new column called hindfoot_cm which contains the hindfoot_length values (currently in mm) converted to centimeters. In this hindfoot_cm column, there are no NAs and all values are less than 3.\

#### View the head() of the data.frame. 

##### Hint: Think about how the commands should be ordered to produce this data frame! 

```{r}
#Q9 

```

#### Q10: How many animals were caught in each plot_type surveyed?

```{r}
#Q10

```

#### Q11: Use group_by() and summarize() to find the mean, min, and max hindfoot length for each species (using species_id). Also add the number of observations (hint: see ?n).

```{r}
#Q11

```

#### Q12: What was the heaviest animal measured in each year? Return the columns year, genus, species_id, and weight.

```{r}
#Q12

```

#### Q13: Reshape the surveys data frame with year as columns, plot_id as rows, and the number of genera per plot as the values. You will need to summarize before reshaping, and use the function n_distinct() to get the number of unique genera within a particular chunk of data. It’s a powerful function! See ?n_distinct for more.

```{r}
#Q13

```

#### Q14: Now take that data frame and pivot_longer() it, so each row is a unique plot_id by year combination.

```{r}
#Q14

```





