---
title: "TidyverseDay2"
author: "YOUR NAME"
output: html_document
date: "2024-02-06"
---

Chunk 0
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Let's start! --------------------------------

Chunk 1
```{r}
#Load libraries
library(tidyverse)
```

Chunk 2
```{r}
#Load data
surveys <- read_csv("diversity.csv")
```

## Dplyr: Selecting columns --------------------------------

Chunk 3
```{r}
#Select
select(surveys, plot_id, species_id, weight)
```

Chunk 4
```{r}
#Select all columns except certain ones 
select(surveys, -record_id, -species_id)
```

## Dplyr: Filtering rows --------------------------------

Chunk 5
```{r}
#Filter
filter(surveys, year == 1995)
```

## Dplyr: Pipes --------------------------------

## Pipes Example #1

Chunk 6
```{r}
#Pipes step by step
#Original data
surveys
```

Chunk 7
```{r}
#Pipes step by step
#Filter
surveys2 <- filter(surveys, weight < 5) 
surveys2
```

Chunk 8
```{r}
#Pipes step by step
#Select
surveys_sml <- select(surveys2, species_id, sex, weight)
surveys_sml
```

Chunk 9
```{r}
#Selecting and filtering using pipes
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```

## Pipes Example #2

Chunk 10
```{r}
#Pipes step by step
#Original data
surveys
```

Chunk 11
```{r}
#Pipes step by step
#Filter
surveys2 <- filter(surveys, genus == "Peromyscus", species == "maniculatus") 
surveys2
```

Chunk 12
```{r}
#Pipes step by step
#Select
surveys_sml <- select(surveys2, sex, genus, species)
surveys_sml
```

Chunk 13
```{r}
#Write the pipe
surveys %>%
```

## Pipes continued

Chunk 14
```{r}
#Piping into new object
surveys_sml <- surveys %>%
  filter(genus == "Peromyscus", species == "maniculatus") %>% 
  select(sex, genus, species)
```

Chunk 15
```{r}
#View new object
surveys_sml
```

## Dplyr: Mutate --------------------------------

Chunk 16
```{r}
#New column of weight in kg
surveys %>%
  mutate(weight_kg = weight / 1000)
```

Chunk 17
```{r}
#Second new column based on the first new column
surveys %>%
  mutate(weight_kg = weight / 1000,
         weight_lb = weight_kg * 2.2)
```

Chunk 18
```{r}
#New column of weight in kg and view head
surveys %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
```

Chunk 19
```{r}
#Remove NAs from weight column
surveys %>%
  filter(!is.na(weight)) %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
```

## Dplyr: Group_by and summarize --------------------------------

# Group_by Example #1

Chunk 20
```{r}
#Pipes step by step
#Original data
surveys
```

Chunk 21
```{r}
#Pipes step by step
#Group by
surveys2 <- group_by(surveys, sex)
surveys2
```

Chunk 22
```{r}
#Pipes step by step
#Select
surveys_sml <- summarize(surveys2, mean_weight = mean(weight, na.rm = TRUE))
surveys_sml
```

Chunk 23
```{r}
#Group_by and summarize
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

## Group_by and summarized continued

Chunk 24
```{r}
#Group_by multiple columns
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>%
  tail()
```

Chunk 25
```{r}
#Omit NAs in weight
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight))
```

Chunk 26
```{r}
#Summarizing multiple variables
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(min_weight)
```

Chunk 27
```{r}
#Arranging in descending order 
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(desc(mean_weight))
```

## Dplyr: Counting --------------------------------

Chunk 28
```{r}
#Count
surveys %>%
	count(sex)
```

Chunk 29
```{r}
#Count a combination of factors
surveys %>%
	count(sex, species)
```

Chunk 30
```{r}
#Count a combination and arrange table in alphabetical order of species and descending order of the count
surveys %>%
  count(sex, species) %>%
  arrange(species, desc(n))
```

## Tidyr: Pivoting --------------------------------

Chunk 31
```{r}
#Filter out NAs in weight, group_by plot_id and genus, summarize mean weight in new column
surveys_gw <- surveys %>%
  filter(!is.na(weight)) %>%
  group_by(plot_id, genus) %>%
  summarize(mean_weight = mean(weight))

surveys_gw
```

Chunk 32
```{r}
#Pivot from long format to wide format
surveys_wide <- surveys_gw %>%
  pivot_wider(names_from = genus, values_from = mean_weight)

surveys_wide
```

Chunk 33
```{r}
#Pivot from wide format to long format
surveys_long <- surveys_wide %>%
  pivot_longer(names_to = "genus", values_to = "mean_weight", cols = -plot_id)

surveys_long
```



