---
title: "TidyverseDay1"
author: "YOUR NAME"
output: pdf_document
date: "2024-02-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Let's start! --------------------------------

Chunk 1
```{r}
#Load libraries
library(tidyverse)
```

Chunk 2
```{r}
#Load data
surveys <- read_csv("diversity.csv")
```

Chunk 3
```{r}
#Extract first few lines of data using head() function
head(surveys)
```

## Inspecting a dataframe --------------------------------
Chunk 4
```{r}
#Inspect the structure of the data frame
str(surveys)
```

Chunk 5
```{r}
#Inspect the names of the data frame
names(surveys)
```

Chunk 6
```{r}
#Summarize the data frame
summary(surveys)
```

## Indexing a dataframe --------------------------------

Chunk 7
```{r}
#First row and column from surveys
surveys[1, 1] 
```

Chunk 8
```{r}
#First row, sixth column
surveys[1, 6]
```

Chunk 9
```{r}
#Select all columns for the first row
surveys[1, ] 
```

Chunk 10
```{r}
#Select the first column across all rows
surveys[, 1]
```

Chunk 11
```{r}
#Select the first three rows of the 5th and 6th column
surveys[c(1, 2, 3), c(5, 6)] 
```

Chunk 12
```{r}
#We can use the : operator to create those vectors for us
surveys[1:3, 5:6] 
```

Chunk 13
```{r}
#This is equivalent to head_surveys <- head(surveys)
head_surveys <- surveys[1:6, ] 
```

Chunk 14
```{r}
#Call column name directly
surveys["species_id"] 
```

## Subsetting a dataframe --------------------------------

Chunk 15
```{r}
#The whole data frame, except the first column
surveys[, -1]    
```

Chunk 16
```{r}
#Equivalent to head(surveys)
surveys[-(7:nrow(surveys)), ] 
```

## Data Types: Conversion to a factor --------------------------------

Chunk 17
```{r}
#Creating a factor
surveys$sex <- factor(surveys$sex)
```

Chunk 18
```{r}
#Check factor conversion
str(surveys)
```

Chunk 19
```{r}
#Summarize factor
summary(surveys$sex)
```

## Data Types: Finding levels --------------------------------

Chunk 20
```{r}
#New value ("sex")
sex <- factor(c("male", "female", "female", "male"))
```

Chunk 21
```{r}
#Visualize levels (order) of data
#Default is alphabetical order
levels(sex)
```

Chunk 22
```{r}
#Number of levels
nlevels(sex)
```

## Data Types: Changing factor order --------------------------------

Chunk 23
```{r}
#Changing order of levels
sex <- factor(sex, levels = c("male", "female"))
```

Chunk 24
```{r}
#Check new order
sex
```


## Data Types: Renaming factors --------------------------------

Chunk 25
```{r}
#Plot
plot(surveys$sex)
```

Chunk 26
```{r}
#Assign variable
sex <- surveys$sex
```

Chunk 27
```{r}
#Add a factor level
sex <- addNA(sex)
```

Chunk 28
```{r}
#Visualize levels (order) of data
levels(sex)
```

Chunk 29
```{r}
#Head variable
head(sex)
```

Chunk 30
```{r}
#Rename third level
levels(sex)[3] <- "undetermined"
```

Chunk 31
```{r}
#Re-plot with new level
plot(sex)
```


## Questions ----------------------------------

# Q1: Based on the output of str(survey) in chunk 4:
1. What is the class of the object surveys? 
2. How many rows and how many columns are in the object?

# Q2: Based on code shown in chunk 9 and 16:
1. Create a data.frame (surveys_200) containing only the data in row 200 of the surveys dataset.

2. Notice how nrow() gave you the number of rows in a data.frame?

- Use that number to pull out just that last row from the surveys dataset.
- Compare that with what you see as the last row using tail() to make sure it’s meeting expectations.
- Pull out that last row using nrow() instead of the row number.
- Create a new data frame (surveys_last) from that last row.

# Q3: Based on code shown in chunk 17:
1. Change the columns taxa and genus in the surveys data frame into a factor.

2. Using the functions you learned before, can you find out…

- How many rabbits were observed?
- How many different genera are in the genus column?

# Q4: Based on code shown in chunk 28:
1. Rename "F" and "M" to "female" and "male" respectively. 

2. Can you recreate the barplot such that "undetermined" is first (before "female")?



