---
title: "Getting Data and Databases (Part 1)"
author: "Jessica Hill"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE)
```

---

## Importing/Exporting data in R

#### <ins> Considerations: How is the data formatted? <ins>

Before importing data into R, we need to check how our data is formatted. Is our data tidy?

Does your data file have row and column names?

- If not, consider adding them. 

- If so, are they using good or bad naming conventions? 
        
```
# Helpful tips

- Avoid names with blank spaces. 
    Good column names: Long_jump or Long.jump
    Bad column name: Long jump

- Avoid names with special symbols: ?, $, *, +, #, (, ), -, /, }, {, |, >, <
    Only underscore can be used.

- Avoid beginning variable names with a number. Use letter instead. 
    Good column names: sport_100m or x100m 
    Bad column name: 100m

- Each variable has its own column. Each row is an observation.
    All data in a column has the same type (e.g. it is all numerical data, or it is all text data)

- Column names must be unique and describe the variable contained within that column. Duplicated names are not allowed.

- No blank columns or blank rows between data

- R is case sensitive. This means that Name is different from Name or NAME.

- Delete any comments in your file

- Replace missing values by NA (for not available)

- If you have a column containing date, use the four digit format. 
    Good format: 01/01/2016 
    Bad format: 01/01/16
    
- Data are consistent (e.g. if a binary variable can take values ‘Yes’ or ‘No’ then only these two values are allowed, with no alternatives such as ‘Y’ and ‘N’)
```

For more helpful tips, see: 

- [Anderson Chapter 2, Module 2.3: The "Tidy" Data Format](https://geanders.github.io/improve_repro/2.3-module3.html#module3)
    
- [Formatting Problems](https://datacarpentry.org/spreadsheet-ecology-lesson/02-common-mistakes/)

```
# Key take aways

- Avoid using multiple tables within one spreadsheet.
- Avoid spreading data across multiple tabs.
- Record zeros as zeros. Dont leave them blank.
- Use an appropriate null value to record missing data.
- Don’t use formatting to convey information or to make your spreadsheet look pretty.
- Place comments in a separate column.
- Record units in column headers.
- Include only one piece of information in a cell.
- Avoid spaces, numbers and special characters in column headers.
- Avoid special characters in your data.
- Record metadata in a separate plain text file.
```

---

#### <ins> Considerations: What is the data type? <ins>

We also need to make sure that the data is saved in a file format that R can recognize. 

What type of file formats can R recognize?

- Quite a lot actually. 
-  Here are some resources to check out below 
    - [Import, Export, and Convert Data Files](https://cran.r-project.org/web/packages/rio/vignettes/rio.html)
    - [R Data Import/Export](https://cran.r-project.org/doc/manuals/r-release/R-data.html) 

The file formats you will most likely work with are .csv, .tsv or .txt, .xlsx, Database, XML/HTML 

---

#### <ins> How can we import data into R? <ins>

Many ways to import data into R, but usually we use a function that is somewhat specific to the file type we are trying to import. 

Lets check out some examples below: 

1)  Read in a .csv file, using 'read.csv'
```{r}
data1 <- read.csv('./Example_Data/heart_data.csv')
head(data1, 5)
```

2)  Now read in the same .csv file, but using 'read.table'
```{r}
data2 <- read.table('./Example_Data/heart_data.csv', sep=",", header = 1)
head(data2, 5)
```

3) Read in a .txt file, using 'read.delim' 
```{r}
data3 <- read.delim("./Example_Data/TulsaJun20_2020.txt",header = F)
head(data3, 7) # if you read in over 7 lines, you will see his whole speech
```

4) Read in an Excel file
```{r}
# Uncomment to install, then recomment when done installing
#install.packages("tidyverse") 
library(tidyverse)
#install.packages("readxl")
library(readxl) 


Data <- read_excel("./Example_Data/HNP_Stats.xlsx", sheet = 1)
head(Data, 5)

Country <- read_excel("./Example_Data/HNP_Stats.xlsx", sheet = 2)
head(Country, 5)

Series <- read_excel("./Example_Data/HNP_Stats.xlsx", sheet = 3)
head(Series, 5)
```

5) Import data from a Database using SQL
```{r}
# Uncomment to install, then recomment when done installing
#install.packages("DBI")
#install.packages("RSQLite")
library(DBI)
library(RSQLite)


# load the database using dbConnect()
conn <- dbConnect(RSQLite::SQLite(), "./Example_Data/basketball.sqlite")

# display the names of the tables
dbListTables(conn)

# run a query and display the results, using the dbGetQuery() function. Just add a SQLite connection object and SQL query as a string.  
print(head(dbGetQuery(conn, "SELECT * FROM player"), 5))

data4 = dbGetQuery(conn, "SELECT * FROM inactive_players")
head(data4, 5)
```

6) Import data from XML files

XML (eXtensible Markup Language) provides a standardized way to represent and store data. It also facilitates data exchange and interoperability between softwares and databases. Finally, it easily allows for web integration. 

Akin to XML, is JSON (JavaScript Object Notation), another data interchange format

[XML Repository with example data](https://www.w3schools.com/xml/)

```{r}
# Uncomment to install, then recomment when done installing
#install.packages("XML", "xml2")
library(XML)
library(xml2)

elt2 = read_xml('./Example_Data/celegans_elt2.xml')
elt2_parsed = xmlParse(elt2)
Dbtag = getNodeSet(elt2_parsed, "//Dbtag")
elt2_df = xmlToDataFrame(elt2_parsed, nodes=Dbtag)
print(elt2_df)
```

7) Read in a JSON (JavaScript Object Notation) file
```{r}
# Uncomment to install, then recomment when done installing
#install.packages("rjson")
library(rjson)

JsonData <- fromJSON(file = './Example_Data/iris.json')
print(JsonData[1:10])

# to convert the JSON data into an R dataframe, use 'as.data.frame()' function in the data.table package
#install.packages("data.table") 
library(data.table)

data6 = do.call("rbind", JsonData)
data6
```

8) Import data from HTML files
```{r}
# Uncomment to install, then recomment when done installing
#install.packages("rvest")
library(rvest)


url <- read_html("https://en.wikipedia.org/wiki/COVID-19_pandemic_deaths#:~:text=There%20have%20been%20reported%207%2C026%2C534,16%20and%2028.2%20million%20deaths.")
tables <- html_nodes(url, "table")
data7 <- html_table(tables[10])
data7 <- as.data.frame(data7)
print(head(data7, 5))
```

```{r}
url2 <- read_html("https://en.wikipedia.org/wiki/List_of_U.S._states_by_life_expectancy")
tables2 <- html_nodes(url2, "table")
data8 <- html_table(tables2[1])
data8 <- as.data.frame(data8)
print(head(data8, 5))
```

---

#### <ins> How can we export data from R? <ins>

There are a several ways to export data from R

But first we need to read in some data, manipulate it, and then we will export it.
```{r}
winequality_df <- read.csv("./Example_Data/winequality-red.csv", header = T)

new_winequality_df <- winequality_df %>% filter(quality > 6)
```

Export a dataframe to a .txt file, using write.table 
```{r}
write.table(new_winequality_df, file = './Example_Data/new_winequality_df.txt', col.names = TRUE,
             row.names = FALSE, sep = "\t")
```

Export a dataframe to a .csv file, using write.table
```{r}
write.table(new_winequality_df, file = './Example_Data/new_winequality_df.csv', col.names = TRUE,
             row.names = FALSE, sep = ",")
```

Now export the same dataframe to a .csv file, using write.csv 
```{r}
write.csv(new_winequality_df, file = './Example_Data/new_winequality_df2.csv', row.names = FALSE)
```

An alternative to write.table (meant to be faster)
```{r}
# Uncomment to install, then recomment when done installing 
#install.packages("data.table")
library(data.table)


fwrite(new_winequality_df, file = './Example_Data/new_winequality_df2.txt', sep = "\t")

fwrite(new_winequality_df, file = './Example_Data/new_winequality_df3.csv')
```

---

#### <ins> Are all data file formats handled equally? <ins>

Lets check the reading speeds of the different functions 
```{r}
# Use system.time() to time how fast R processes an expression. 
# This function is especially useful for comparing the speed of two coding options.

system.time(
  read.csv("./Example_Data/winequality-red.csv", header = T)
)
```

```{r}
system.time({
  read.table("./Example_Data/winequality-red.csv", header = T, sep = ",");
  #read.delim("./Example_Data/new_winequality_df.txt", header = T, sep = "\t")
})
```

What about writing speeds? 
```{r}
system.time({
  write.csv(new_winequality_df, file = './Example_Data/new_winequality_df2.csv', row.names = FALSE)
})
```

```{r}
system.time({
fwrite(new_winequality_df, file = './Example_Data/new_winequality_df3.csv') 
})
```

---

[Now lets move to Part 2](GettingData_Databases_Part2.md)
